---
title: "Notes on Statistics"
output: html_document
---

#Part II: Modeling

```{r, echo=FALSE}
set.seed(1)
```

```{r, echo=FALSE}
#rotate <- function(x) t(apply(x, 2, rev))
rotate <- function(x) t(x[nrow(x):1,])
draw.matrix <- function(x, g=TRUE) {
  image(rotate(x), axes=FALSE, col = grey(seq(1, 0, length = 256)),
         xlab=paste("matrix", nrow(x), "x", ncol(x)))
  box()
  if(g)grid(nx=ncol(x), ny=nrow(x), col="gray")
}
```

```{r}
head(cars)
(n <- nrow(cars))
x <- cars$speed
y <- cars$dist
```

##Linear Model

Equation: $Y = X \hat{\beta} + \epsilon$

$$\left[\begin{array}
{r}
Y_1 \\
... \\
Y_N
\end{array}\right]
 = \left[\begin{array}
{rrrr}
1 & x_1 & ... & x_{p-1} \\
... & ... & ... & ... \\
1 & x_{1,N} & ... & x_{p-1,N}
\end{array}\right]
\left[\begin{array}
{r}
\beta_0 \\
\beta_1 \\
... \\
\beta_{p-1} 
\end{array}\right]
 + \left[\begin{array}
{r}
\epsilon_0 \\
... \\
\epsilon_N 
\end{array}\right]
$$

X: N x p ( rows x columns)

Solution (LSE)

$$X^T Y = X^T X \hat{\beta}$$

$$ \hat{\beta} = (X^T X)^{-1} X^T Y $$

If $X = QR, Q^TQ=I$, then $R\hat{\beta} = Q^TY$ (back-substitution)

Hat-matrix

$$ \hat{Y} = H Y $$

$$ \hat{Y} = X \hat{\beta} $$

$$ H = X (X^T X) ^ {-1} X^T $$

Mean Square Error

$$ r = Y - \hat{Y} = (I - H) Y $$
$$ MSE = \frac{1}{N - p} r^Tr $$

Weighted LSE, best $C=W^T W$ is the inverse of $Y$ covariance matrix

$$ \hat{\beta} = (X^T W ^ T W X)^{-1} X^T W^T W Y $$


Sample: Cars

```{r}
X <- cbind(1, x)
head(X)
X <- model.matrix(~x)
head(X)
(b <- solve(crossprod(X)) %*% crossprod(X, y))

h <- X %*% solve(crossprod(X)) %*% t(X)
dim(h) # Hii is dinstance from mean (centroid) of X
sum(diag(h)) # model d.f.

op <- par(mfrow = c(1, 2))
matplot(1:n, h, type="l")
points(diag(h), pch=17, col="red")
draw.matrix(h, g=F)
par(op)

#QR optimization
QR <- qr(X)
Q <- qr.Q(QR)
R <- qr.R(QR)
backsolve(R, crossprod(Q, y))
solve.qr(QR, y) #QR
#solve(R, crossprod(Q,Y))

fitted <- X %*% b # fit existing, or new X
head(fitted, 3)
head(tcrossprod(Q)%*%y, 3) #QR
residuals <- y - fitted
sum(residuals)

# standardized residuals
zresiduals <- (residuals - mean(residuals)) / sd(residuals) # scale(residuals)
sresiduals <- residuals / (sd(residuals) * (1 - diag(h)) ^ 2)

(mse <- crossprod(residuals, residuals) / (n - ncol(X)) )

op <- par(mfrow = c(3, 2))
plot(x, y, col="gray")
abline(h=mean(y), lty=2, col="gray")
lines(x, fitted, col="blue", type="l", lwd=2)
plot(x, fitted, col="blue", type="l", lwd=2)
abline(h=mean(y), lty=2, col="gray")
points(x, residuals + mean(y), col=ifelse(residuals > 0, "red", "green"), pch=ifelse(residuals > 0, 6, 2))
plot(fitted, residuals, col=ifelse(residuals > 0, "red", "green"))
abline(h=0, lty=2, col="gray")
plot(diag(h), type="h")
abline(h=2 * mean(diag(h)), lty=2, col="gray")
plot(diag(h), zresiduals, col=ifelse(residuals > 0, "red", "green"), ylab="zresiduals")
plot(diag(h), sresiduals, col=ifelse(residuals > 0, "red", "green"), ylab="sresiduals")
par(op)
```

Confounding

```{r}
# confounding present if different (collinear)
cat("ncol =", ncol(X), "rank =", qr(X)$rank)
```

Model

```{r}
#formula(dist ~ speed)
fit <- lm(dist ~ speed, data=cars)
summary(fit) # coef(fit)
c(fit$rank, fit$qr$rank)

op <- par(mfrow = c(1, 2))
plot(x, fit$fitted.values, col="blue")
abline(a=fit$coefficients[1], b=fit$coefficients[2], lty=2, col="red")
#abline(fit)
plot(x, fit$residuals)
abline(h=0, lty=2, col="gray")
par(op)
```

Model diagnostics

```{r}
op <- par(mfrow = c(2, 2))
plot(fit)
par(op)
```

```{r}
op <- par(mfrow = c(3, 3))
plot(dffits(fit)) # how x affects fitted
matplot(dfbetas(fit), type="l") # how betas change if x is exlcuded (sd relative)
plot(cooks.distance(fit), type="h") # effect of deleting a given observation
plot(hatvalues(fit), type="h") # leverage
plot(rstandard(fit))
plot(rstudent(fit))
plot(covratio(fit))
plot(density(rstudent(fit)), main="")
par(op)

head(influence.measures(fit)$infmat, 1)
cn <- colnames(influence.measures(fit)$infmat)
matplot(influence.measures(fit)$infmat, type="l", ylim=c(-1,1.1)) # summary
legend("bottomleft", legend=cn, lty=1:length(cn), col=1:length(cn), inset=c(-0.1,-0.1), xpd=TRUE)

car::outlierTest(fit)

op <- par(mfrow = c(1, 3))
car::qqPlot(fit, simulate=T, labels=row.names(fit))
#influence plot
#influence = leverage (= hatvalues) * discrepency (= dfbetas, or cook's D)
plot(hatvalues(fit), rstudent(fit), type="n", xlab="hatvalues", ylab="sresiduals") # set up
cook <- cooks.distance(fit)
points(hatvalues(fit), rstudent(fit), cex=10 * cook / max(cook), pch=21, bg="gray") # zoom 10
abline(v=2 * mean(diag(h)), h=c(-2,0,2), lty=2) # interesting hatvalues, and sresiduals
# join influence (added-variable (partial-regression) plots) - of no use here, given cars$speed only
car::avPlots(fit)
par(op)


```

Prediction

```{r}
xnew = c(16, 18, 23)
as.vector(cbind(1, xnew) %*% b)
predict(fit, list(speed=xnew))
```

Confidence

```{r}
confint(fit)
prd <- predict(fit, list(speed=x), interval = c("confidence"), level = 0.95, type="response")
plot(x, fit$fitted.values, col="blue", type="b", main="Confidence 95%")
points(x, y, col="gray")
#lines(x, prd[,2],col="red",lty=2)
#lines(x, prd[,3],col="red",lty=2)
matlines(x, prd[,2:3],col="red",lty=2)
```

Analysis of variance

```{r}
(a<- anova(fit))
# speed reduces SS by 
a[["Sum Sq"]][1]
# p-value
1 - pf(a[["F value"]][1], a[["Df"]][1], a[["Df"]][2])
```

```{r}
# autocorrelation between residuals (idependent error terms)?
car::durbinWatsonTest(fit)
```

Comparison

```{r}
fit2 <- lm(dist ~ speed + I(speed^2), data=cars)
anova(fit2, fit)

# Akaike's information criterion (AIC) - lack of fit
AIC(fit2, fit) # smaller is better

speed.factor <- factor(cars$speed)
anova(lm(cars$dist ~ cars$speed + speed.factor))
```

```{r}
#http://www.statmethods.net/stats/rdiagnostics.html
car::vif(fit2)
sqrt(car::vif(fit2)) > 2 # multi-collinear
car::ncvTest(fit) #non-constant error variance test
#car::spreadLevelPlot(fit)
```


##Estimation Errors

Model std. errors

```{r}
summary(fit)$coe[,2]
```

Standard error of an estimate is the standard deviation of the sampling distribution of an estimate

```{r}
b <- replicate(1000, {
  index <- sample(n, 30)
  scars <- cars[index,]
  x <- scars$speed
  y <- scars$dist
  b <- lm(y ~ x)$coef
  b
})
b <- t(b)
head(b, 1)
apply(b, 2, sd)

op <- par(mfrow = c(2, 2))
hist(b[,1], main="intercept")
qqnorm(b[,1])
qqline(b[,1])
hist(b[,2], main="speed")
qqnorm(b[,2])
qqline(b[,2])
par(op)
```

Correlation of estimates

```{r}
cor(b[,1], b[,2])
```

Exact variance of estimates:

$$
\mbox{var}(\mathbf{AY}) = \mathbf{A}\mbox{var}(\mathbf{Y}) \mathbf{A}^\top 
$$

$$\mbox{var}(\boldsymbol{\hat{\beta}}) = \sigma^2\mathbf{(X^\top X)^{-1}}$$

$$
\mathbf{r}\equiv\boldsymbol{\hat{\varepsilon}} = \mathbf{Y}-\mathbf{X}\boldsymbol{\hat{\beta}}$$

$$ \Sigma = s^2 \equiv \hat{\sigma}^2 = \frac{1}{N-p}\mathbf{r}^\top\mathbf{r} = \frac{1}{N-p}\sum_{i=1}^N r_i^2$$

```{r}
p <- ncol(X)
residuals <- y - fitted # r
XtXinv <- solve(crossprod(X))
s <- sqrt(sum(residuals^2) / (n - p)) # rss = sum(residuals^2) = crossprod(residuals)
s
sqrt(crossprod(residuals) / (n - p)) #same s
(ses <- sqrt(diag(XtXinv)) * s) # same std. errors as lm
```

QR

```{r}
df <- length(y) - QR$rank
sigma2 <- sum(residuals^2)/df
beta.var <- sigma2 * chol2inv(qr.R(QR))
sqrt(diag(beta.var))
```

t-test
```{r}
(tstat <- summary(fit)$coe[,1] / summary(fit)$coe[,2])
2 * (1 - pt(abs(tstat), n - 2))
```

## Models

Let fake some hypothetical data, cars with wonder friction (0 control) and speed class:

```{r}
n <- nrow(cars)
wcars <- cars
wcars[["odist"]] <- cars$dist # remeber original speed
# wunder friction 1, makes stop distance shorter
wcars[["fric"]] <- as.factor(rep(paste(sep="", "f", 0:1), length.out = n))
index.f1 <- wcars$fric == 'f1'
wcars[index.f1, "dist"] <- round(wcars[index.f1, "dist"] - 0.3 * wcars[index.f1, "dist"])
# speed class: very slow, slow, fast, very fast
mu <- mean(wcars$speed)
ss <- sd(wcars$speed)
wcars[wcars$speed < (mu - ss), "sclass"] <- "VS"
wcars[(wcars$speed >= (mu - ss)) & (wcars$speed < mu), "sclass"] <- "S"
wcars[(wcars$speed >= mu)  & (wcars$speed < (mu + ss)), "sclass"] <- "F"
wcars[wcars$speed >= (mu + ss), "sclass"] <- "VF"
wcars$sclass <- factor(wcars$sclass, levels=c("VS", "S", "F", "VF"))
wcars$group <- factor(paste0(wcars$fric, "." , wcars$sclass))
head(wcars)
tail(wcars)
```

```{r}
op <- par(mfrow = c(2, 2))
stripchart(wcars$dist ~ wcars$fric, data=wcars, vertical=TRUE, method="jitter", col=wcars$fric)
stripchart(wcars$dist ~ wcars$sclass, data=wcars, vertical=TRUE, method="jitter", col=1:length(levels(wcars$sclass)))
boxplot(wcars$dist ~ wcars$fric, ylab="stop distance", xlab="friction type")
boxplot(wcars$dist ~ wcars$sclass, ylab="stop distance", xlab="speed class")
par(op)
op <- par(mfrow = c(1, 2))
plot(wcars$speed, wcars$dist, col=wcars$fric, main="Cars by friction")
plot(wcars$speed, wcars$dist, col=wcars$sclass, main="Cars by speed class")
par(op)
```

1. Single factor parameters (one parameter per factor level - 1)

```{r}
model <- model.matrix(~ fric, data=wcars) # relevel if needed
head(model)
draw.matrix(model) # custom function based on image

# changing reference level
levels(wcars$fric)
(c <- contrasts(wcars$fric))
contrasts(wcars$fric) <- contr.treatment(levels(wcars$fric), base=2)
contrasts(wcars$fric)
contrasts(wcars$fric) <- c # reset back
contrasts(wcars$fric)

cat("ncol =", ncol(model ), "rank =", qr(model )$rank) # confounding if different

fit <- lm(dist ~ fric, data=wcars)
summary(fit)$coef
anova(fit) # one way anova
s <- split(wcars$dist, wcars$fric)
mean(s[["f0"]]) # fit$coef[1]
mean(s[["f0"]]) - mean(s[["f1"]]) # fit$coef[2]

stripchart(dist ~ fric, data=wcars, vertical=TRUE, method="jitter", col=c(1,2))
abline(h=c(fit$coef[1], fit$coef[1] + fit$coef[2]), col=c("black", "red"), lty=2)
legend("topright", legend=c("intercept (f0)", "f1"), col=c(1,2), lty=2)

plot(effects::allEffects(fit), multiline=TRUE)

t.test(s[["f1"]], s[["f0"]], var.equal=TRUE)$statistic # same as from lm
```


2. Combined factors

```{r}
model <- model.matrix(~ fric + sclass, data=wcars)
head(model)
draw.matrix(model)
contrasts(wcars$sclass)

fit <- lm(dist ~ fric + sclass, data=wcars)
summary(fit)$coef
cc <- c("black", "red", "blue", "green", "magenta")
stripchart(split(wcars$dist, wcars$group), 
           vertical=TRUE, pch=1, method="jitter", col=cc, main="Coeficients")
abline(h=c(fit$coef[1],
           fit$coef[1] + fit$coef[2],
           fit$coef[1] + fit$coef[2] + fit$coef[3],
           fit$coef[1] + fit$coef[2] + fit$coef[3] + fit$coef[4],
           fit$coef[1] + fit$coef[2] + fit$coef[3] + fit$coef[4] + fit$coef[5]), 
       col=cc, lty=2)

plot(effects::allEffects(fit), multiline=TRUE)
summary(effects::allEffects(fit))
```

Coeficients are weighted means of the groups (weighted by group size)

3. Factor interaction

```{r}
# ~ wcars$fric + wcars$sclass + wcars$fric : wcars$sclass
model <- model.matrix(~ fric * sclass, data=wcars)
colnames(model)
draw.matrix(model)
cat("ncol =", ncol(model ), "rank =", qr(model )$rank) # confounding if different
fit <- lm(dist ~ fric * sclass, data=wcars)
summary(fit)$coef
anova(fit) # two way anova
plot(wcars$speed, fit$fitted.values, col=wcars$group)
plot.design(dist ~ fric * sclass, data=wcars)

plot(effects::allEffects(fit), multiline=TRUE)
```

Quasi-standard errors

```{r}
fit$contrasts
library(qvcalc)
qvcalc(fit, "sclass")
```

No intercept

```{r}
model <- model.matrix(~ 0 + group, data=wcars)
head(model)
draw.matrix(model)
fit <- lm(dist ~ 0 + group, data=wcars)
summary(fit)$coef
sapply(split(wcars$dist, wcars$group), mean)
plot(wcars$speed, fit$fitted.values, col=wcars$group)
# no anova
```

4. Numeric parameters

```{r}
# ~ wcars$speed + wcars$fric + wcars$speed : wcars$fric
model <- model.matrix(~ speed * fric, data=wcars)
colnames(model)
draw.matrix(model)
cat("ncol =", ncol(model ), "rank =", qr(model )$rank) # confounding if different
fit <- lm(dist ~ speed * fric, data=wcars)
summary(fit)$coef
anova(fit) # ancova - analysis of covariance 
```

Two lines

```{r}
b <- fit$coef
y0 <- b["(Intercept)"] + b["speed"] * wcars$speed
y1 <- (b["(Intercept)"] + b["fricf1"]) + (b["speed"] + b["speed:fricf1"]) * wcars$speed

plot(wcars$speed, wcars$dist, col="gray")
lines(wcars$speed, y0, type="b", col="red")
lines(wcars$speed, y1, type="b", col="blue")
legend("topleft", legend=c("f0", "f1"), col=c("red", "blue"), lty=1)

plot(effects::allEffects(fit), multiline=TRUE)
```


Model simplification

```{r}
step(fit)
```


## Contrast

$C\hat{\beta}$ - linear combination $C$ over $\hat{\beta}$, a 0 in $C$ means no involvement of that $\hat{\beta}$ coeficient. Error of $C$: $\sqrt{C \Sigma C^T}$

```{r}
fit <- lm(dist ~ fric * sclass, data=wcars)
summary(fit)$coef
```

```{r}
suppressMessages(library(contrast))
slowf0.vs.f1 <- contrast(fit,
                   list(sclass="VS", fric = "f0"), 
                   list(sclass="VS", fric = "f1"))
slowf0.vs.f1
```

```{r}
#install.packages("multcomp")
suppressMessages(library(multcomp))
# fricf1:sclassF - fricf1:sclassS
C <- matrix(c(0,0,0,0,0,-1,1,0), 1)
interaction <- glht(fit, linfct=C)
summary(interaction)
fit$coef[7] - fit$coef[6]
```

All differences of differences between parameters are given by ANOVA:

```{r}
#aov(dist ~ fric * sclass, data=wcars)
anova(fit)
```


##Smoothing - Local Weighted Regression (loess)

```{r}
fit <- loess(y~x, degree=1, span=1/3)
newx <- seq(min(y), max(x), len=100) 
smooth <- predict(fit, newdata=data.frame(x=newx))
plot(x, y, col="darkgrey", pch=16)
lines(newx, smooth, col="black", lwd=3)
lines(lowess(y~x, f=1/3), col="red")
legend("topleft", legend=c("loess", "lowess"), col=c("black", "red"), lty=1)
#car::qqPlot(residuals(loess(y~x)))
```

## Predicting with Linear Regression

```{r}
dataIdx <- (1:nrow(cars) %% 2 == 0) #mod
testIdx <- (1:nrow(cars) %% 2 == 1)
sum((dataIdx - testIdx)) #0
Y <- cars[dataIdx,]$dist
X <- cars[dataIdx,]$speed
#cross validation
test.Y <- cars[testIdx,]$dist
test.X <- cars[testIdx,]$speed
c(length(X), length(test.X))
c(mean(Y), mean(test.Y))
c(sd(Y), sd(test.Y))

plot(X, Y, col="black", pch=16)
points(test.X, test.Y, col="darkgray", pch=16)
```

```{r}
fit <- lm(Y ~ X)
yhat <- round(predict(fit))
test.yhat <- round(predict(fit, newdata=data.frame(X=test.X)))
# 1 - E(Y∣X=x) (conditional probability)
#error:
1 - mean(abs(Y - yhat) < 5)
1 - mean(abs(test.Y - test.yhat) < 5)
1 - mean(abs(test.Y - yhat) < 5)
plot(X, yhat, type="b")
points(test.X, test.yhat, pch=16, col="darkgrey")
```

## Predicting with K-nearest Neighbor (KNN)

```{r}
library(class)
k=3
yhat <- as.numeric(knn(data.frame(X), data.frame(X), Y, k=k))
test.yhat <- as.numeric(knn(data.frame(X), data.frame(test.X), Y, k=k))
#error
1 - mean(abs(Y - yhat) < 5)
1 - mean(abs(test.Y - test.yhat) < 5)
1 - mean(abs(test.Y - yhat) < 5)

plot(X, yhat, type="b")
points(test.X, test.yhat, pch=16, col="darkgrey")
```

```{r}
train.error <- rep(0, 16)
test.error <- rep(0, 16)
idx <- seq(along=train.error)
for(k in idx){
  yhat <- knn(data.frame(X), data.frame(X), Y, k=k)
  train.error[k] <- 1 - mean((as.numeric(yhat) - 1) == y)
  
  yhat <- knn(data.frame(X), data.frame(test.X), Y, k = k)
  test.error[k] <- 1 - mean((as.numeric(yhat) - 1) == y)
}
plot(idx, train.error, type="n", xlab="K", ylab="error")
lines(idx, train.error,type="b", col="black", lty=2, lwd=2)
lines(idx, test.error,type="b", col="darkgray", lty=3, lwd=2)
legend("bottomright", c("Train","Test"), col=c("black","darkgray"), lty=c(2,3), box.lwd=0)
```

##Maximum-Likelihood Estimation (MLE)

MLE is a generalized form of least-squares estimation (LSE).

<http://cran.r-project.org/doc/contrib/Ricci-distributions-en.pdf>

```{r}
library(stats4)
```

Mean and standard deviation

```{r}
z <- (cars$speed - mean(cars$speed)) / sd(cars$speed) # standartize speed
c(mean(z), sd(z))

# estimate mu, sigma using log-likelihood function (L)
LL <- function(mu, sigma) {
  R <- dnorm(z, mu, sigma) 
  -sum(log(R))
}
(fit <- mle(LL, start = list(mu = 0, sigma=1)))

#difference
coef(fit) - c(mean(z), sd(z))

# speed (no standartization)
c(mean(cars$speed), sd(cars$speed))
LL <- function(mu, sigma) {
  R <- dnorm(cars$speed, mu, sigma) 
  -sum(log(R))
}
(fit <- mle(LL, start = list(mu = 14, sigma=3),
            method = "L-BFGS-B", lower = c(-Inf, 0),
            upper = c(Inf, Inf)))

#difference
coef(fit) - c(mean(cars$speed), sd(cars$speed))

```

Linear models

```{r}
#install.packages("bbmle")
library(bbmle)
fit <- lm(dist ~ speed, data=cars)
summary(fit)$coef
residuals <- cars$dist - fit$fitted.values
c(mean(residuals), sd(residuals))

LL <- function(beta0, beta1, mu, sigma) {
  R <- cars$dist - cars$speed * beta1 - beta0
  R <- suppressWarnings(dnorm(R, mu, sigma))
  -sum(log(R))
}

fit <- mle2(LL, start=list(beta0=-15, beta1=3, mu=0, sigma=5))
coef(fit)

residuals <- cars$dist - cars$speed * coef(fit)["beta1"] - coef(fit)["beta0"]
c(mean(residuals), sd(residuals))
```

## Logistic Regression

Binary outcome $logit(p) = X \beta$ (no projection error). [Deviance](http://en.wikipedia.org/wiki/Deviance_(statistics)) changes are $\chi^2$ distributed (d.f. = change in number of parameters)

```{r}
cars$fastcar <- as.factor(ifelse(cars$dist > mean(cars$dist), 1, 0))
cars$fastcar
```

```{r}
fit <- glm(fastcar~speed,family=binomial("logit"), data=cars)
exp(cbind(coef(fit), confint(fit))) # odds scale

summary(fit)
anova(fit, test="Chisq")

op <- par(mfrow = c(1, 2))
plot(cars$speed, fitted(fit), type="l")
xx <- seq(0, 1, 0.01)
plot(xx, log(xx/ (1 - xx)), type="l")
par(op)

(p <- predict(fit, newdata = list(speed=c(10,20)), type = "response"))
ifelse(p > 0.5, "Fast", "Slow")

op <- par(mfrow = c(2, 2))
plot(fit)

par(op)
```

Overdispersion

```{r}
c(fit$deviance, fit$df.residual, summary(fit)$dispersion)
(fi <- deviance(fit) / df.residual(fit)) # overdispersion if too > 1

# if fi >> 1
qfit <- glm(fastcar~speed,family=quasibinomial("logit"), data=cars)

c(qfit$deviance, 
  qfit$df.residual,
  summary(qfit)$dispersion,
  summary(fit)$dispersion * fit$df.residual, 
  summary(qfit)$dispersion * qfit$df.residual,
  deviance(fit),
  deviance(qfit)
  )

pchisq(summary(qfit)$dispersion * fit$df.residual, fit$df.residual, lower = F) # p-value < 0.05 for overdispersion

anova(fit, qfit, test="Chisq")
```

## Robust Regression

```{r}
fitLse <- lm(dist ~ speed, data=cars)
```

1. Least Absolute Values (LAV), uses median (higher estimation variance that LSE)

```{r}
suppressMessages(suppressWarnings(library(quantreg)))
fitLav <- rq(dist ~ speed, data=cars)
summary(fitLav)
```

2. M-Estimation: 2a) Huber (combines: lse around center, lav in sides), 2b) Bisquare weights (MM) (resistant to extreme observations in tails)

```{r}
library(MASS)
fitHuber <- rlm(dist ~ speed, data=cars)
fitBisquare <- rlm(dist ~ speed, data=cars, method="MM") # or robustbase::lmrob
summary(fitHuber)
summary(fitBisquare)
```

3. Least Trimmed Squares (LTS), uses ordered 50% lower LSE, robust, needs more data

```{r}
suppressMessages(suppressWarnings(library(robustbase)))
fitLts <- ltsReg(dist ~ speed, data=cars)
summary(fitLts)
```

4. Least Median Squares (LMS)

```{r}
#library(MASS)
fitLms <- lqs(dist ~ speed, data=cars)
coef(fitLms)
```

```{r}
models <- list(lse=fitLse, lav=fitLav, huber=fitHuber, bisq=fitBisquare, lts=fitLts, lms=fitLms)
cmodels <- matrix(0, length(models), 2)
rownames(cmodels) <- names(models)
colnames(cmodels) <- c("intercept", "speed")
plot(cars$speed, cars$dist, col="gray")
for(i in 1:length(models)) {
  cc <- coef(models[[i]])
  cmodels[i, 1] <- cc[1]
  cmodels[i, 2] <- cc[2]
  abline(coef=cc, col=i, lty=i)
}
abline(h=c(median(cars$dist), mean(cars$dist)),
       v=c(median(cars$speed), mean(cars$speed)),
       lty=c(1,2), col=8)
legend("topleft", legend=names(models),
       col=c(1:length(models), 8,8), lty=c(1:length(models), 8,8))

cmodels

# weights (huber gives more 1, lse is all 1)
plot(fitBisquare$w, fitHuber$w, xlab="bisquare weights", ylab="huber weights", xlim=c(0,1), ylim=c(0,1))
abline(0,1)
```

```{r}
for(i in 1:length(models)) {
  cc <- coef(models[[i]])
   print(cc)
} 
```


##Hierachical Models

Base average (population)

$$\theta \approx N(\mu, \tau^2) $$

Sample given base

$$Y|\theta \approx N(\theta, \sigma^2)$$

$$E(\theta | Y) = B\mu + (1 - B) Y$$

Shrink sample towards population average

$$B = \frac{ \sigma^2}{ \sigma^2 + \tau^2 }$$


**EOF**
