---
title: "R Language Quirks"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Enviroment

*Environments consist of a frame, or collection of named objects, and a pointer to an enclosing environment.*

Default environment frame is `R_GlobalEnv`. Top one is `R_EmptyEnv` that cannot have parents or data.

```{r}
#?environment
environment()
globalenv()
environmentName(environment())

search()

e <- environment()
while(environmentName(e) != "R_EmptyEnv") {
  print(environmentName(e))
  e <- parent.env(e)
}
print(environmentName(e))
```

Functions get their own environment frame, blocks not.

```{r}
f <- function() {
  print("new frame")
  e <- environment()
  print(e)
  print(parent.env(e))
  print(parent.frame())
  if(TRUE) {
    print("same frame")
    print(environment())
  }
  {
    print("same frame")
    print(environment())
  }
  f2 <- function() {
    print("another new frame")
    print(environment())
    print(parent.env(environment()))
  }
  f2()
}
f()
environment(f);
```

Parent enviroment scope is lexical:

```{r}
f <- function() {
  print("same parent")
  e <- environment()
  print(e)
  print(parent.env(e))
  print("frames")
}

f2 <- function() {
  f()
  f3 <- function () {
    print("nested")
    f()
  }
  f3()
}
f2()

```

This affect closures:

```{r}
v <- "A";
f <- function() {
  print(v)
}
f2 <- function() {
  print("prints A:")
  f();
  v <- "C"
  print("prints A again:")
  f();
}


f()
f2()

v <- "B"
f()
print("now prints A is really B")
f2()

```

Hence, currying in best done using nested functions:

```{r}
a <- 1
b <- 2
bind <- function(a, b) {
  function(x) {
    (a + b) * x
  }
}

bound1 <- bind(2,3);
bound1(1)
bound2 <- bind(3, 3);
bound2(1)
bound1(1)
```


The `<<-` operator will set or create (top) a parent frame variable:

```{r}
f <- function() {
  a <- "a"
  b <<- "b"
}
f()
ls()
rm(list = ls())
ls()
```

Environments can be created:

```{r}
e <- new.env()
print(e)
print(parent.env(e))
assign('a', 1, envir=e)
exists('a', envir = e)
exists('a')
get('a', envir = e)
#?with # ads a temp top frame from data
```

##Lazy Evaluation

```{r}
action <- function(a) { print(a); return(a) }

ifelse(TRUE, action(1), action(2))
```

Function arguments are evalued first when used:

```{r}
gmean <- function(x, p, n = length(x), w = rep(1/n, n)) 
  as.vector(if(p == 0) prod(sapply(1:n, function(i) { x[i] ^ w[i] })) ^ (1 / sum(w)) 
  else (crossprod(w, x ^ p) / sum(w)) ^ (1 / p))

gmean(1:10, 0) #geometric mean
gmean(1:10, 1) #arithmetic mean
```

Not used vars are ignored:

```{r}
rm(list = ls())
f <- function(a, b) {
  print(a)
}
f(3)
```

`switch` is therefore a function:

```{r}
action <- function(a) { print(a); return(a) }
switch(2, action(1), action(2), action(3))
```

##Return is a Function

```{r}
f <- function() {
  return(3 + ifelse(TRUE, return(sqrt(4)), return(sqrt(9))))
}
f()
```


##Operators

Binary operators are infix functions:

```{r}
2 * 3 - 1
`-`(`*`(2, 3), 1)
```

Custom operators:

```{r}
`%aha%` <- function(l, r) {
  rep(l, r)
}

2 %aha% 3

`%aha%`(2, 3)

1 %aha% 2 %aha% 3
```

##Data Types

Vectors (of constants):

```{r}
(a <- 5)
length(a)
dim(a)
names(a)
names(a) <- c('A') 
names(a)
str(a)
(a <- 1:15)
dim(a) <- c(3, 5)
dim(a)
nrow(a)
ncol(a)
typeof(a)
# now this one makes a list!
(a <- c(function(){ print("a") }, function(){ print("b") }))
a[[1]]()
typeof(a)
```

Pairwise vector operators

```{r}
x <- 1:10
y <- 2:11
x + y
```

##Attributes

```{r}
a <- 1:5
str(a)
attr(a, "aha") <- "vvv"
attr(a, "aha")

(a <- 1:15)
dim(a) <- c(3, 5)
a
dim(a)
attributes(a)
attributes(a) <- NULL
dim(a)
a

m <- matrix(1:15, nrow=3, ncol=5)
class(m)
attr(m, "class")
attributes(m)
```


##Objects and Classes

* http://www.programiz.com/r-programming/object-class-introduction
* http://adv-r.had.co.nz/OO-essentials.html

###S3 Objects

S3 objects are list objects with `class` attribute set:

```{r}
p <- list(x=3, y=4)
class(p) <- "point"
class(p)
attr(p, "class")
p$x
p$y
```

A S3 constructor function:

```{r}
point <- function(x, y) {
  o <- list(x=x, y=y)
  attr(o, "class") <- "point"
  o
}

p <- point(3, 4)
class(p)
attr(p, "class")
```

Methods:

```{r}
#methods(print)
#methods(class="default")

print(p) #no specialization

print.point <- function(p) {
  print(paste(p$x, p$y))
}

print(p)
```

Custom generic methods:

```{r}
aha <- function(o) {
  UseMethod("aha")
}

aha.point <- function(p) {
  print("aha point")
}

aha.default <- function(p) {
  print("aha default")
}

aha(p)
a <- 3
aha(3)
p
#print.aha
#summary.aha
#plot.aha

plot.aha <- function(p, ...) {
  plot(p$x, p$y, ...)
}
plot(p, pch=20, col=4)
```

###S4 classes

```{r}
point <- setClass("point", slots=list(x="numeric", y="numeric"))
p <- new("point", x=3, y=4)
p
attributes(p)
str(p)
isS4(p)
p@x
p@y
p <- point(x=3, y=5)
p@y
p@y <- 6
slot(p, "y") <- 6
show(p)
```

S4 methods:

```{r}
#showMethods()
isS4(show)
#showMethods(show)

setMethod("show", "point", function(object) {
  print(paste(object@x, object@y))
})

show(p)
```

###Reference class (RC)

```{r}
rpoint <- setRefClass("rpoint", 
          fields=list(x="numeric", y="numeric"),
          methods=list(
            add = function(other) {
              x <<- other + x
              y <<- other + y
            },
            move = function(k) {
              x <<- k * x
              y <<- k * y
            },
            show = function() {
              print(paste(x, y))
            }
          ))

p <- rpoint(x=4, y=5)
show(p)
p$move(3)
p$x
p$y
```

RC objects are by reference:

```{r}
p <- point(x=3, y=5)
rp <-rpoint(x=4, y=5)

modify <- function (p) {
  p@x <- 6
}
rmodify <- function (p) {
  p$x <- 6
}

p
modify(p)
p

rp
rmodify(rp)
rp
```

**EOF**



